SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization\\$\;$\\DRAFT

### 2013-03-22 paper/paper.tex
paper updates.

\begin{abstract}
QuickCheck, developed by Klassen and Hughes, is a powerful library for automatic
test-case generation.  Because QuickCheck performs random testing, it can
discover very large counterexamples.  QuickCheck provides an interface for the
user to write \emph{shrink} functions to attempt to reduce the size of
counterexamples discovered.  Hand-written implementations of shrink can be
difficult to write: they may be complex, inefficient, and consist of significant
boilerplate code.  We introduce \emph{SmartCheck}, an automatic efficient
counterexample reduction algorithm.  SmartCheck reduces algebraic data using
generic search heuristics to efficiently find smaller counterexamples
independent of the program or property being tested.  In addition to shrinking,
SmartCheck generalizes counterexamples by returning a \emph{formula} that
abstracts over individual counterexamples and make it easier for the programmer
to uncover multiple counterexamples at once.  SmartCheck has been implemented
for Haskell and is freely available.
\end{abstract}

@@

Typically, effectively shrinking complex data requires a good understanding of
how shrinking in QuickCheck works and the semantics of the property and program
being evaluated.  Bad definitions can be so inefficient that they are unusable.

@@

To illustrate, consider the example in Figure~\ref{fig:initial}.\footnote{All
  examples and algorithms in this paper are presented in
  Haskell~\cite{haskell98}.  We strive to use idiomatic Haskell, and only rely
  on the language extensions of existential type quantification and scoped type
  variables.}  Data type \ttp{T} is a product type containing four lists of wrapped
\ttp{Int16}s and a single wrapped \ttp{Word8}.  For our purposes, it is not important
what \ttp{T} models, but for concreteness, suppose it models the values in four
linked lists in an operating system, together with an 8-bit status register.

Now suppose we are modeling some program that serializes values of type \ttp{T}.
The input to the program satisfies the invariant \ttp{pre}, that the sum of
values in each list of \ttp{Int16}s is less than or equal to 256 (and by
definition, the \ttp{Word8} value is less than or equal to 256).  Assuming this,
we want to show \ttp{post} holds, that the sum of all the values from \ttp{T} is
less than $5 * 256$, where five is the number of fields in \ttp{T}.  At first
glance, the property seems reasonable.  But we have forgotten about underflow;
for example, the value
%
\begin{code}
T (N 10) [N (-20000)] [N (-20000)] [] []
\end{code}
%
\noindent
satisfies \ttp{pre} but fails \ttp{post} (\ttp{==>} is implication from the
QuickCheck library).

@@

A more clever programmer that understands the semantics of QuickCheck's
\ttp{shrink} implementation,\footnote{The following approach was suggested to
  the author by John Hughes.} defines a \ttp{shrink} instance as follows:
%
\begin{code}
shrink (T w i0 i1 i2 i3) = map go xs
  where xs = shrink (w, i0, i1, i2, i3)
        go (w', i0', i1', i2', i3') =
          T w' i0' i1' i2' i3'
\end{code}
%
\noindent
This ``tuple shrink'' definition does not suffer the same shortcomings: it
shrinks a value even if it contains an empty list, and the combinatorial blowup
of shrink candidates is avoided, since a pair \ttp{(a,b)} is shrunk by
attempting to shrink \ttp{a} while holding \ttp{b} constant, then attempting to
shrink \ttp{b} while holding \ttp{a} constant (and is generalized to larger
tuples).  Still, using this definition, from some initial counterexamples,
shrinking may take over 15 seconds and result in a final counterexample
containing nearly 60 \ttp{N} constructors.\footnote{All results reported in the paper
  are with a GHC-compiled program, using \ttp{-O2}, on a 4-core 2.7GHz Intel~i7
  processor.}  We would like to do better, particularly since the data type and
property are so simple.

@@

Finally, even if we could overcome the problems described above with QuickCheck
and SmallCheck, we are left with two others.
\begin{itemize}
  \item A counterexample is usually representative of other counterexamples, but
    that generalization is left to the user to discover for herself.  For
    example, consider the counterexample presented earlier:
%
\begin{code}
T (N 10) [N (-20000)] [N (-20000)] [] []
\end{code}
%
\noindent
For any \ttp{Word8}, the counterexample still holds.  Thus, we can be more
informative and return a quantified formula rather than a single counterexample:
%
\begin{code}
forall x . T x [N (-20000)] [N (-20000)] [] []
\end{code}
%
\noindent
denoting that for any \ttp{Word8}, the counterexample holds.  Such a formula has
the added benefit of replacing a potentially large sub-term of a composite data
type value with a quantifier, reducing the sheer size of the term.

  \item Second, during testing, it is sometimes helpful to cover the input space
    by discovering different counterexamples to \ttp{prop}.  In the example in
    Figure~\ref{fig:initial}, one counterexample is fairly representative of
    others, but this is not true in general (e.g., consider properties about
    abstract syntax trees).  Ideally, the testing framework would automatically
    add a predicate characterizing a counterexample as the precondition to
    \ttp{prop} that strengthens the property.  That is, for a specific
    counterexample \ttp{ex}, we have a predicate \ttp{predEx} that holds if its
    argument is another counterexample that is similar to (i.e., some
    generalization of) \ttp{ex}.  Then we test against a new property
%
\begin{code}
prop' a = not (cex a) ==> prop a
\end{code}
%
%% \noindent
%% where \ttp{==>} is QuickCheck's implication operator that discards tests that
%% violate the precondition.
\end{itemize}

@@

Motivated by these limitations of QuickCheck and SmallCheck, we developed
\emph{SmartCheck}.  SmartCheck takes a counterexample produced by some
oracle---in these experiments, we use QuickCheck---and generically minimizes and
generalizes the counterexample, as described in the preceding bullet points.

@@

And we stress that SmartCheck achieves these results \emph{generically},
requiring no shrink definitions to be provided by the user.  Indeed, there is
nothing particularly special about the example given.

\paragraph{Contributions}
This paper, and the library it describes, makes the following contributions:

\begin{itemize}

\item an efficient counterexample reduction strategy for algebraic
  data types (Section~\ref{sec:shrinking});

\item an approach to counterexample generalization, to present the user with a
  formula describing a set of counterexamples to a property
  (Section~\ref{sec:generalization});

\item an approach to generically strengthen a property with a precondition that
  characterizes an already-discovered counterexample for the purpose of finding
  new, dissimilar counterexamples (Section~\ref{sec:generalization});

\item an implementation based on generic programming that is not only more
  efficient that hand-written \ttp{shrink} instances in many cases but that
  generates the instances automatically for the user
  (Section~\ref{sec:implementation}).

%% \item More generally, this paper is the first to explore the idea of
%%   efficient, generic counterexample reduction and generalization for
%%   functional program testing.
\end{itemize}

\noindent
In Section~\ref{sec:experiments}, we discuss some of our experiences with using
SmartCheck, including checking properties from XMonad, and related work and
concluding remarks are made in Sections~\ref{sec:related}
and~\ref{sec:conclusions}, respectively.

@@

\subsection{Existential Sub-Value Generalization}

@@

XMonad passes all of its QuickCheck tests, but let us see what might happen to a
new contributor if things go awry.  Suppose a developer defines a deletion
function to delete a window, if it exists.  There is a deletion function that
exists in XMonad, which is quite complex, given the amount of state that is managed
by \ttp{StackSet}.  However, one function used in deletion is to filter the
stack of windows associated with each workspace defined:
%
\begin{code}
removeFromWorkspace ws =
  ws { stack = stack ws >>= filter (/=w) }
\end{code}
%
\noindent
Now, suppose the programmer makes a simple typo (or perhaps has a more serious
misunderstanding) and instead writes
%
\begin{code}
removeFromWorkspace ws =
  ws { stack = stack ws >>= filter (==w) }
\end{code}
%
\noindent
When testing the property \ttp{prop\_delete}, which says that deleting the
focused window of the current stack removes it from the \ttp{StackSet} \ttp{x}.
%
\begin{code}
prop_delete x =
    case peek x of
        Nothing -> True
        Just i  -> not (member i (delete i x))
\end{code}
%
\noindent
QuickCheck returns
%
\medskip%
\begin{sloppypar}
\scriptsize
\noindent%
\ttp{StackSet \{current = Screen \{workspace = Workspace \{tag = NonNegative
      \{getNonNegative = 0\}, layout = -1, stack = Just (Stack \{focus = 'S', up =
        "", down = ""\})\}, screen = 1, screenDetail = 1\}, visible = [Screen
    \{workspace = Workspace \{tag = NonNegative \{getNonNegative = 2\}, layout = -1,
        stack = Nothing\}, screen = 2, screenDetail = -1\},Screen \{workspace =
      Workspace \{tag = NonNegative \{getNonNegative = 3\}, layout = -1, stack =
        Nothing\}, screen = 0, screenDetail = -1\}], hidden = [Workspace \{tag =
      NonNegative \{getNonNegative = 1\}, layout = -1, stack = Just (Stack \{focus
        = `\char`\\NUL', up = "", down = ""\})\},Workspace \{tag = NonNegative
      \{getNonNegative = 4\}, layout = -1, stack = Just (Stack \{focus = 'I', up =
        "", down = ""\})\}], floating = fromList []\} }
\end{sloppypar}
\medskip%
\noindent
This is a relatively small counterexample, but even the smallest \ttp{StackSet}
values are somewhat large due to the number of fields within it.  But let us see
if we can do better.

@@

\section*{Acknowledgments}

>>>>

\begin{abstract}
QuickCheck, developed by Claessen and Hughes, is a powerful library for automatic
test-case generation.  Because QuickCheck performs random testing, some of the
counterexamples discovered are very large.  QuickCheck provides an interface for
the user to write \emph{shrink} functions to attempt to reduce the size of
counterexamples discovered.  Hand-written implementations of shrink can be
difficult to write: they may be complex, inefficient, and consist of significant
boilerplate code.  We introduce \emph{SmartCheck}, an automatic efficient
counterexample reduction algorithm.  SmartCheck reduces algebraic data using
generic search heuristics to efficiently find smaller counterexamples
independent of the program or property being tested.  In addition to shrinking,
SmartCheck generalizes counterexamples by returning a \emph{formula} that
abstracts over individual counterexamples and make it easier for the programmer
to uncover multiple counterexamples at once.  SmartCheck has been implemented
for Haskell and is freely available.
\end{abstract}

@@

Defining an efficient and effective shrink method requires a good understanding
of how shrinking in QuickCheck works and the semantics of the property and
program being evaluated.  Bad definitions can be so inefficient that they are
unusable.

In addition, shrinking is one side of the coin when it comes to making counterexamples more
understandable.  The other side is generalization.  For example, the following
is a relatively small counterexample returned when QuickChecking an (incorrect)
implementation in Xmonad, a popular X11 window manager written in
Haskell~\cite{xmonad}:
%
\medskip%
\begin{sloppypar}
%\footnotesize
\noindent%
\ttp{StackSet \{current = Screen \{workspace = Workspace \{tag = NonNegative
      \{getNonNegative = 0\}, layout = -1, stack = Just (Stack \{focus = 'S', up =
        "", down = ""\})\}, screen = 1, screenDetail = 1\}, visible = [Screen
    \{workspace = Workspace \{tag = NonNegative \{getNonNegative = 2\}, layout = -1,
        stack = Nothing\}, screen = 2, screenDetail = -1\},Screen \{workspace =
      Workspace \{tag = NonNegative \{getNonNegative = 3\}, layout = -1, stack =
        Nothing\}, screen = 0, screenDetail = -1\}], hidden = [Workspace \{tag =
      NonNegative \{getNonNegative = 1\}, layout = -1, stack = Just (Stack \{focus
        = `\char`\\NUL', up = "", down = ""\})\},Workspace \{tag = NonNegative
      \{getNonNegative = 4\}, layout = -1, stack = Just (Stack \{focus = 'I', up =
        "", down = ""\})\}], floating = fromList []\} }
\end{sloppypar}
\medskip%
\noindent
Functional programmers should be familiar with having to debug a ``wall of
text'' when showing data structures.  What if instead a counterexample like the
following were returned by the testing framework:
%
\begin{code}
forall values x0 x1 x2 x3:
  StackSet
    (Screen (Workspace x0 (-1) (Just x1)) 1 1)
    x2 x3 (fromList [])
\end{code}
%
\noindent
The formula quantifies away all the irrelevant portions of the data structure
with respect to the counterexample, meaning that for any well-typed values that
replace \ttp{x0}, \ttp{x1}, \ttp{x2}, and \ttp{x3}, the counterexample still
holds.

Motivated by the problems of reducing and generalizing large counterexamples, we
have developed \emph{SmartCheck}.  SmartCheck takes a counterexample produced by
some oracle---in these experiments, we use QuickCheck---and generically
minimizes and generalizes the counterexample.

As demonstrated by a motivating example in Section~\ref{sec:example}, SmartCheck
can shrink more than an order-of-magnitude faster than efficient hand-written
shrink implementations for some properties and programs.  More importantly, the
time required to shrink scales linearly with the size of the
counterexample---other common shrink implementations often scale in polynomial
time or worse.  And the counterexamples returned are two to five times smaller.

After presenting some preliminary definitions in
Section~\ref{sec:preliminaries}, in Section~\ref{sec:shrinking}, we describe the
generic shrinking algorithm implemented.  The algorithm is particularly novel in
that it generates new random sub-values during the reduction phase (see related
work in Section~\ref{sec:related}).

%%  than the tuple or list-truncation approaches
%% described earlier, and the size of resulting counterexamples are two-five times
%% smaller (more detailed performance numbers are given in
%% Section~\ref{sec:performance}).  Furthermore, other shrink approaches tend to
%% have ``long tails,'' meaning that for some original counterexamples, that are
%% not shrunk much or take a very long time to shrink.
%% Motivated by these limitations of QuickCheck and SmallCheck, we developed
%% \emph{SmartCheck}.

SmartCheck implements three novel approaches to automatically generalize
counterexamples, which are described in Section~\ref{sec:generalization}.  The
first algorithm universally quantifies sub-values that always fail in tests, as
briefly described earlier in the introduction.  The second algorithm
existentially quantifies sub-values for types in which every possible variant
fails the property.  For example, finding counterexamples (\ttp{Left 2}) and
(\ttp{Right True}) for the type (\ttp{Either Int Bool}) means there exists a
counterexample regardless of the variant chosen.

The third algorithm automatically strengthens properties by omitting ``similar''
counterexamples to the ones previously observed.  The algorithm is motivated by
noting that there are often multiple ways in which a property may fail; for
example, a property that parsing a pretty-printed abstract syntax tree (AST)
results in the original AST may fail due to multiple bugs.  During testing, it
is useful to discover all these possibilities in one go.  In practice, the
problem is solved by discovering a counterexample \ttp{cex}, abstracting it, and
then adding a new precondition to the property that informally says ``omit
counterexamples of form \ttp{cex}.''  Adding preconditions manually is laborious
and may cause the programmer to make premature fixes to the program, if she
believes she has isolated the error.

We describe our implementation based on generic programming in
Section~\ref{sec:implementation}; the implementation is open-source.  In
Section~\ref{sec:experiments}, we discuss some of our experiences with using
SmartCheck, including checking properties from XMonad and a natural language
processing library.

\section{A Motivating Example}\label{sec:example}

@@

In this section, we motivate in more detail the challenges in shrinking
counterexamples (we focus on shrinking and only touch on generalization here).
Consider the example in Figure~\ref{fig:initial}.\footnote{All examples and
  algorithms in this paper are presented in Haskell 2010~\cite{haskell2010} plus
  the language extensions of existential type quantification and scoped type
  variables.}  Data type \ttp{T} is a product type containing four lists of
wrapped \ttp{Int16}s and a single wrapped \ttp{Word8}.  For our purposes, it is
not important what \ttp{T} models, but for concreteness, suppose it models the
values in four linked lists in an operating system, together with an 8-bit
status register.

Now suppose we are modeling some program that serializes values of type \ttp{T}.
The input to the program satisfies the invariant \ttp{pre}, that the sum of
values in each list of \ttp{Int16}s is less than or equal to 256 (and by
definition, the \ttp{Word8} value is less than or equal to 256).  Assuming this,
we want to show \ttp{post} holds, that the sum of all the values from \ttp{T} is
less than $5 * 256$, where five is the number of fields in \ttp{T}.  At first
glance, the property seems reasonable.  But we have forgotten about underflow;
for example, the value
%
\begin{code}
T (N 10) [N (-20000)] [N (-20000)] [] []
\end{code}
%
\noindent
satisfies \ttp{pre} but fails \ttp{post} (the \ttp{==>}) operator in the figure
implements implication from the QuickCheck library).

@@

A more clever programmer that understands the semantics of QuickCheck's
\ttp{shrink} implementation,\footnote{The following approach was suggested to
  the author by John Hughes.} defines a \ttp{shrink} instance as follows:
%
\begin{code}
shrink (T w i0 i1 i2 i3) = map go xs
  where xs = shrink (w, i0, i1, i2, i3)
        go (w', i0', i1', i2', i3') =
          T w' i0' i1' i2' i3'
\end{code}
%
\noindent
This ``tuple shrink'' definition does not suffer the same shortcomings: it
shrinks a value even if it contains an empty list, and the combinatorial blowup
of shrink candidates is avoided, since a pair \ttp{(a,b)} is shrunk by
attempting to shrink \ttp{a} while holding \ttp{b} constant, then attempting to
shrink \ttp{b} while holding \ttp{a} constant (and is generalized to larger
tuples).  Thus, each argument to \ttp{T} is independently shrunk.

Still, using this definition, from some initial counterexamples, shrinking may
take over 15 seconds and result in a final counterexample containing nearly 60
\ttp{N} constructors.\footnote{All results reported in the paper are with a
  GHC-compiled program, using \ttp{-O2}, on a 4-core 2.7GHz Intel~i7 processor.}
We would like to do better, particularly since the data type and property are so
simple.

@@

%% Finally, even if we could overcome the problems described above with QuickCheck
%% and SmallCheck, we are left with two others.
%

%% Second, during testing, it is sometimes helpful to cover the input space
%%     by discovering different counterexamples to \ttp{prop}.  In the example in
%%     Figure~\ref{fig:initial}, one counterexample is fairly representative of
%%     others, but this is not true in general (e.g., consider properties about
%%     abstract syntax trees).  Ideally, the testing framework would automatically
%%     add a predicate characterizing a counterexample as the precondition to
%%     \ttp{prop} that strengthens the property.  That is, for a specific
%%     counterexample \ttp{ex}, we have a predicate \ttp{predEx} that holds if its
%%     argument is another counterexample that is similar to (i.e., some
%%     generalization of) \ttp{ex}.  Then we test against a new property
%% %
%% \begin{code}
%% prop' a = not (cex a) ==> prop a
%% \end{code}
%
%% \noindent
%% where \ttp{==>} is QuickCheck's implication operator that discards tests that
%% violate the precondition.
%\end{itemize}

Here we provide some illustrative comparisons between SmartCheck's shrinking
against other shrinking approaches.  As a representative example, we use the
example presented in Section~\ref{sec:intro}.  The results for SmartCheck are
fairly generalizable to other types and properties since SmartCheck's reduction
algorithm has no dependencies on the specifics of either; however, user-defined
shrink methods may vary significantly based on the property and counterexample
type.

@@

@@

Finally, note that after shrinking the counterexample, SmartCheck will also
generalize it.  Note that for any \ttp{Word8} value in the first argument of
\ttp{T}, the counterexample still holds.  Thus, we can be more informative and
return a quantified formula rather than a single counterexample:
%
\begin{code}
forall x . T x [N (-20000)] [N (-20000)] [] []
\end{code}

%% \begin{itemize}

%% \item an efficient counterexample reduction strategy for algebraic
%%   data types (Section~\ref{sec:shrinking});

%\item an approach to counterexample generalization, to present the user with a
%%   formula describing a set of counterexamples to a property
%%   (Section~\ref{sec:generalization});

%% \item an approach to generically strengthen a property with a precondition that
%%   characterizes an already-discovered counterexample for the purpose of finding
%%   new, dissimilar counterexamples (Section~\ref{sec:generalization});

%\item

%% \item More generally, this paper is the first to explore the idea of
%%   efficient, generic counterexample reduction and generalization for
%%   functional program testing.
% \end{itemize}

@@

\subsection{Existential Sub-Value Generalization}\label{sec:existential}

@@

XMonad passes all of its QuickCheck tests, but let us see what might happen to a
new contributor if things go awry.  Suppose a developer defines a deletion
function to delete a window, if it exists.  There is a deletion function that
exists in XMonad, which is quite complex, given the amount of state that is managed
by \ttp{StackSet}.  However, one function used in deletion is to filter the
stack of windows associated with each workspace defined:
%
\begin{code}
removeFromWorkspace ws =
  ws \{ stack = stack ws >>= filter (/=w) \}
\end{code}
%
\noindent
Now, suppose the programmer makes a simple typo (or perhaps has a more serious
misunderstanding) and instead writes
%
\begin{code}
removeFromWorkspace ws =
  ws \{ stack = stack ws >>= filter (==w) \}
\end{code}
%
\noindent
When testing the property \ttp{prop\_delete}, which says that deleting the
focused window of the current stack removes it from the \ttp{StackSet} \ttp{x}.
%
\begin{code}
prop_delete x =
    case peek x of
        Nothing -> True
        Just i  -> not (member i (delete i x))
\end{code}
%
\noindent
%QuickCheck returns
\todo{fill in}

This is a relatively small counterexample, but even the smallest \ttp{StackSet}
values are somewhat large due to the number of fields within it.  But let us see
if we can do better.

@@

\section*{Acknowledgments}
I thank Joe Hurd, John Launchbury, and John Hughes for insightful comments on
earlier drafts of this paper.  In particular, John Hughes discovered errors and
omissions that we have fixed.
